---
title: "cyclone follow up with dittoSeq"
author: "Daniel Bunis"
date: "2023-07-17"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
suppressMessages({
  library(SingleCellExperiment)
  library(dittoSeq)
  library(BiocStyle)
  library(cyclone)
})
```

# Overview

After you have completed running the cyclone pipeline, there will be a number of
useful outputs, including:

- 'batch_qc_plots.pdf'
- 'feature_plots.png'
- 'plots.pdf'
- 'split_umap_by_clustering.png'
- various checkpoint_<#>.RData files.

There are many directions that one can follow up with just these outputs.

- Batch effect assessment: 'batch_qc_plots.pdf' contains 2 umap plots (one colored by and one faceted by the "pool_id" batch information from your file_metadata) as well as numerous heatmaps that either show data by batch directly, or are annotated with batch information. These plots can be useful for assessing if clusters or other features of your data heavily associate with processing batch.  Depending on your experimental design, such association may be expected, but when unexpectedly present, it can be a good idea to perform some form of batch correction. `cyCombine` and `CytoNorm` are two pathways for batch correction of CyTOF data which our team has used.

- Cluster Annotation: Now that your similar cells are grouped together into "clusters", it's time to figure out what each cluster represents. This can be a pretty manual process, but is one of the most important steps for making biological sense of the data. It lets us make sense of what differences in cluster frequencies (see below) between samples actually mean. The 'split_umap_by_clustering.png', 'feature_plots.png', and 'plots.pdf' outputs are all useful here. 'split_umap_by_clustering.png' and 'feature_plots.png' can be used together to determine cell type / lineage defining markers that each cluster expresses, or for a different view, the heatmaps contained within 'plots.pdf' show the median expression of each marker within each cluster.

There are also many other directions that your follow up analysis might take you -- too may for us to attempt to build them all in to cyclone directly.  This vignette will focus some of the most common follow up directions, as well as how to turn your cyclone outputs into a `SingleCellExperiment` object in order to interface with other tools such as the visualization package `dittoSeq`. 

## Scope of this vignette

In this vignette, we will walk through a few common follow up directions.
Specifically, we will start with direct cyclone outputs and then work through
these follow up steps:

1. Consolidate the data into a SingleCellExperiment object
2. Add cell type annotations for clusters & additional metadata (clinical, image coordinates, etc.)
3. Explore the data set with dittoSeq (a color vision deficiency & novice coder
friendly visualization tool)
4. Run statistics on differences in cluster and cell type frequencies between
samples.

# 1. Importing into an SCE

Here, we make use of the SingleCellExperiment object structure because it holds
everything we need for our cytometry analysis: space for expression matrices,
per-cell metadata, dimensionality reductions like umap, and even per-marker
metadata although we won't be making use of that in this vignette.

## Digression1: Why make an SCE?

In addition to the fact that it can be nice simply to have all the relevant data
in a single object, for saving, loading, and version management purposes, using
the Bioconductor-standard SCE format eases interfacing with any of the plethora
of single-cell analysis tools which are built around the structure. dittoSeq is
but one of many such tools.

## Digression2: A mini summary of the important SCE components

The 'assays' slot holds matrices in the form of features (rows) by cells
(columns).  They can be accessed with `assay(<sce>, <assay_name>)` OR a default
assay will be retrieved with `assay(<sce>)`.

The 'colData' slot holds metadata for cells, and the 'rowData' slot holds
metadata for markers. Here, we only make use of 'colData' but the slots work
similarly. Their structure is a DataFrame where each rows holds data for a
col (cell) or row (marker) of the object, and columns are the different bits of
information about them. For example, we will have column in 'colData' holding
the 'cluster' assignments of each cell.  For convenience, the SCE maintainers
set up this syntax `<sce>$<colData_column>` to pull directly from colData, which
makes these data quite easy to access!

The 'reducedDims' slot holds dimensionality reductions, such as UMAP, as
matrices. There are accessor functions for these as well, `?reducedDims`.

## Actually loading in our data

We can make use of helper functions included in the `cyclone` package for to
load our primary pipeline outputs both into R and into a SingleCellExperiment.
Specifically, `make_or_load_full_sce()`. The main bits of the function are:

1. Loading checkpoint1 before checkpoint8 because the cell_metadata element of
the former is updated, with cluster mapping, in the latter.
2. Generating the SCE with code relatively equivalent to the lines below:

```
sce <- SingleCellExperiment(
    assays = list(transformed=t(trans_exp)),
    colData = DataFrame(cell_metadata[, !grepl("UMAP",colnames(cell_metadata))]),
    reducedDims = list(umap=cell_metadata[, grepl("UMAP",colnames(cell_metadata))]))
```

See `?make_or_load_full_sce` for more details.

```{r load full object}
full_sce <- make_or_load_full_sce(
    # sce_file_name is not used in the first pass because we don't yet have a
    #  file at this location, but in later passes, this can load the saved
    #  version!
    sce_file_name = "cytof_full.Rds",
    # Make sure you update these next two to point toward where you have your
    #  own cyclone outputs saved!
    checkpoint1 = "checkpoint_1.RData",
    checkpoint8 = "checkpoint_8.RData",
    load_checkpoints = TRUE
)
```

Now let's look at the summary of the object:

```{r display full_sce}
full_sce
```

# 2. Add cell type annotations for clusters & additional metadata

An early step of any cluster-based analysis is making sense of what each
cluster represents. cyclone's direct outputs serve quite well for this purpose:

'split_umap_by_clustering.png' and 'feature_plots.png' can be used together to
determine cell type / lineage defining markers that each cluster expresses, or
for a different view, the heatmaps contained within 'plots.pdf' show the median
expression of each marker within each cluster.

We recommend compiling your cluster assignments in Excel or other
table/spreadsheet manipulation tool, then saving them as a .csv or .tsv. (There
are also tools for loading directly from Excel files, but we won't cover them
here.) As long as you have one column containing the cell ids, and another
holding cell type names, that's enough!  But generally, it's useful to record
both 'coarse' and 'fine'-level annotations.

A great structure to follow for an annotations.csv is something like:

| cluster | coarse | fine |
|---:|:---:|:---:|
| 1 | CD4T | CD4T_naive |
| 2 | NK | NK_mature |
| 3 | Monocyte | Monocyte_classical |

## Read in your annotations

```{r load and view annotation file}
annots <- read.csv("annotations.csv")
head(annots)
```

## Add annotations to the SCE

This process is simple because of R's `factor` function!  Factors are a useful
data structure in R where "levels" (potential values) of the data are
1. pre-defined and 2. ordered.  With the factor function, you can pick a vector
to start with, set the levels and their order with the `levels` input, and then
update what any of those levels are called with the `labels` input. Usefully,
if any levels are given a matching label, they will be combined together. Thus,
we can make use of this single function to achieve everything we need here --
renaming from cluster_ids to annotations & combination of clusters given same
annotations!

We'll make use of that function to create and pull in both depths of annotation
by starting with the cluster metadata / colData.

```{r view clusters}
head(full_sce$cluster)
```

We'll make a "coarse_annot" metadata / colData column for the coarse-level. 

```{r add coarse_annot}
full_sce$coarse_annot <- factor(
    full_sce$cluster,
    levels = annots$cluster,
    labels = annots$coarse
)
```

We'll make a "fine_annot" metadata / colData column for the fine-level.

```{r add fine_annot}
full_sce$fine_annot <- factor(
    full_sce$cluster,
    levels = annots$cluster,
    labels = annots$fine
)
```

## Pull in any additional metadata

Ideally, you will have added any sample-specific metadata into the
`sample_metadata` that went into the cyclone pipeline at the start. Anything
recorded within `sample_metadata` will have been copied over to the
`cell_metadata` in the `checkpoint_8.Rdata` that we used for creating our SCE.
Thus, all that metadata will already be accessible!

But sometimes we receive certain metadata only later. Such data will need to be
mapped and loaded in before use.

Here, we'll go through addition of donor sex information that was missing in the
`sample_metadata` given to cyclone.

```{r load and view sex metadata}
sex_meta <- read.csv("sex_metadata.csv")
head(sex_meta)
```

As you can see, we have 'ID' and 'sex' in this additional metadata file.

Critically, this 'ID' column can be made to match with our SCE's 'donor_id'
column:

```{r view donor_id}
head(full_sce$donor_id)
```

Thus, we can match the 'ID' column to our SCE's 'donor_id' column in order to
map this sex information to all cells of the data set.

```{r add sex metadata}
full_sce$sex <- sex_meta$Sex[
    match(
        # Target order
        paste0("XVIR1-", full_sce$donor_id),
        # Current data order
        sex_meta$ID
    )
]
# View a random few
rand_cells <- sample(1:ncol(full_sce), 6, replace = FALSE)
colData(full_sce)[rand_cells, c("donor_id", "sex")]
```

Depending on the random seed, you may notice that some samples have a sex of
"N/A" because that data is still not known for the control samples.
We'll deal with this later.

## Specific note for imaging data
We plan to create a standard ingestion process for cells' image coordinates in a
future version of the cyclone pipeline.

Until that point, you can use the procedure here to pull in your cell
coordinates.

To do so, you will need a file structured similarly to the 'sex_metadata.csv'
file used above, but with its ID column mapping to to the cell 'id' colData of
your SCE / column of your cell_metadata, and all necessary coordinates
information in remaining columns.

In the code below, we'll target coordinates from a 'coordinates_metadata.csv'
file structured as: 

| ID | x_loc | y_loc | z_stack |
|---:|:---:|:---:|:---:|
| 1 1 | 17 | 116 | 1 |
| 1 2 | 243 | 10 | 1 |
| 1 3 | 197 | 84 | 1 |

```
coord_meta <- read.csv("coordinates_metadata.csv")
index_matches <- match(
    # Target order
    full_sce$id,
    # Current data order
    coord_meta$ID
)
full_sce$x_loc <- sex_meta$x_loc[index_matches]
full_sce$y_loc <- sex_meta$y_loc[index_matches]
full_sce$z_stack <- sex_meta$z_stack[index_matches]
```

Afterwards, you can use dittoScatterPlot (more details in a later section) to
visualize annotations, metadata, or expression by cells' original location!

Example:
```
dittoScatterPlot(
    full_sce,
    x.by = "x_loc",
    y.by = "y_loc",
    color.by = "<what-to-overlay>",
    # Optional selection of a single z-stack
    cells.use = full_sce$z_stack == "<single-stack-id>"
    show.others = FALSE
)
```

## Saving the SCE with annotations and added metadata folded in

Now, we can save our SCE to be able to skip re-running a lot of this code in the
future, using cyclone's `save_sce` helper function. (Alternatively, you can use
R's `save` or `saveRDS` function directly.  This `save_sce` function is just a
simple wrapper on top of `saveRDS` that automatically sets `compress = TRUE` for
SCEs with more than a million cells.)

```{r, eval = FALSE}
save_sce(full_sce, "cytof_full.Rds")
```

# 3. Exploring the data set with dittoSeq
(a color vision deficiency & novice coder friendly visualization tool)

## Make some plots with dittoSeq!

The exact set of plots that will be useful for any given data set depends
heavily upon the experimental design and upon what sample metadata is actually
available. Maybe you won't need some of these plots, maybe you will. Our goal
in this section is to give a relatively comprehensive overview of plots that
are commonly useful. Feel free skip around... none of the code in this section
makes changes to the underlying sce object, so feel free to only run chunks that
seems useful for your own data!

dittoSeq offers plenty of plotting functions that are useful for cytometry data.

- dittoDimPlot
- dittoFreqPlot
- dittoBarPlot
- dittoScatterPlot
- dittoPlot

**Primary/Required Inputs:** All these functions will take in a target 'object'
(our down_sce for testing or full_sce when ready to make the real version), as
well as one or more marker or metadata names to their 'var' and/or 'group.by',
'color.by', and 'sample.by'. Those alone are minimally enough to make a plot.

**Customize-ability:** All of them also have plenty of quite useful tweaks and
added functionality built in as well. We'll go through a few examples, but we'd
encourage you to check out dittoSeq's own vignette and the functions' own
documentation (e.g. `?dittoDimPlot`) to learn more!

### The basic workflow

**1. Create an SCE representing a subset of the data.**

Because it can take multiple minutes to calculate and display even a single plot
from millions of cells, we recommend creating a (further) downsampled version of
your 'full_sce' so that visualizations can be tested out and tweaked more
quickly. Here, we'll give ourselves an object containing just 100k cells using
cyclone's `downsample_sce` helper function.

```{r downsample}
set.seed(42)
down_sce <- downsample_sce(
    full_sce,
    n_keep = 100000)
down_sce
```

**2. Test and tweak plots on the reduced SCE object.** (use 'object = down_sce')

**3. Make final plots using the full SCE object.** (switch to using
'object = full_sce')

**Note: For the final version of this vignette, we will be using the 'full_sce'.
To implement this suggested workflow for your own data, replace the 'full_sce'
in code below with 'down_sce' for the initial views and tweaking stage, then
switch back to 'full_sce' when producing final figures!**

### UMAP Plots with dittoDimPlot

Marker expression or cell metadata can be plotted on the UMAP using
`dittoDimPlot()`

Some use cases:

- Assessing distributions of sample features or experimental arms
- Viewing locations of clusters in the umap space.

Primary inputs = object and var. (You can leave out those input names if you
like, as long as you provide the 'object' first, and 'var' second.)

```{r dittoDimPlot basic}
# CD3 Marker Expression
dittoDimPlot(object = full_sce, var = "CD3")

# Sample metadata example 1: sample groups or processing batch metric
dittoDimPlot(full_sce, "pool_id")

# Sample metadata example 2: coarse-level cell type annotations
dittoDimPlot(full_sce, "coarse_annot")
```

#### Some particlarly useful dittoDimPlot tweaks

```{r dittoDimPlot tweaks}
# Label the color groups & remove the legend
dittoDimPlot(full_sce, "coarse_annot",
    # Add labels
    do.label = TRUE,
    # Remove the legend
    legend.show = FALSE)

# Adjust plotting order, and also the title
dittoDimPlot(full_sce, "CD16",
    # Plot cells with higher expression in the front with 'order = "increasing"'
    #   Also try: "decreasing" or "randomize"
    order = "increasing",
    # Adjust the plot title
    main = "CD16 Expression")

# Only highlighting certain cells with 'cells.use'
dittoDimPlot(full_sce, "cluster",
    cells.use = full_sce$cluster==4)

## Faceting is VERY useful!
# Example 1: Simple recreation of the effect of the 'split_umap_by_clustering.png'
dittoDimPlot(down_sce, "cluster",
    # Create faceted plots where points are 'split' into different facets by a
    #   metadata given to 'split.by'.
    # Faceting can help make distribution differences more visible!
    split.by = "cluster")

# Example 2: Batch Correction Assessment
#   Ideally, every batches would have cells in all the gray regions of the umap
dittoDimPlot(full_sce, "pool_id",
    split.by = "pool_id")
```

### BoxPlots of cluster or cell type frequencies per sample with dittoFreqPlot

Cell frequencies per-sample, grouped by one or more important metadata can be
plotted using `dittoFreqPlot()`

Use case:

- Visualize if/how cell frequencies are different across sample groups within
your data set

Primary inputs = object, var, sample.by, and group.by.
(You can leave out those input names if you like, as long as you provide the
'object' first, and 'var' second.)

**Note:** You can use cyclone's `freq_stats` function to run wilcoxon-based
statistical testing on cell frequencies calculated with `dittoFreqPlot`, and we
provide details on that process in the final section of this vignette!

```{r dittoFreqPlot basic}
# Coarse-level frequencies per sample, between sexes
dittoFreqPlot(object = full_sce, var = "coarse_annot",
              sample.by = "donor_id", group.by = "sex")

# Cluster frequencies per sample, between sexes, with fine-level annotations in
#    facet labels.
dittoFreqPlot(object = full_sce,
              var = paste(full_sce$fine_annot, full_sce$cluster, sep = "__"),
              sample.by = "donor_id", group.by = "sex")
```

#### Some particlarly useful dittoFreqPlot tweaks

```{r dittoFreqPlot tweaks}
# Allow y-axis to shrink/stretch per data in each facet
dittoFreqPlot(full_sce, "coarse_annot", "file_name", group.by = "sex",
    split.adjust = list(scale="free_y"))

# Only show certain cell types
#   Here: all fine-level annotations in coarse-level "CD4T"
dittoFreqPlot(full_sce, "fine_annot", "file_name", group.by = "sex",
    vars.use = unique(full_sce$fine_annot[full_sce$coarse_annot=="CD4T"]))

# **Adjust percentage normalization to a restricted universe** & only show cell
#   types within that universe.
#   Here: all fine-level annotations in coarse-level "CD4T" cells
dittoFreqPlot(full_sce, "fine_annot", "file_name", group.by = "sex",
    # cells.use adjusts the 'universe' for percent calculation.
    cells.use = full_sce$coarse_annot=="CD4T",
    # vars.use limits cells types shown, but does not affect percent calculation
    vars.use = unique(full_sce$fine_annot[full_sce$coarse_annot=="CD4T"]))

# Coarse-level frequencies per sample, between sexes + also between batch
#   subgroup with color.by
# Also targeting just 4 cell types, for visibility in the same sized plot
#   For all cells, you'd want to make this plot quite large!
dittoFreqPlot(full_sce, "coarse_annot", "file_name",
    group.by = "pool_id",
    color.by = "sex",
    vars.use = c("CD8T", "NK", "B", "ASC"))
```

### Stacked Bar Plots of cluster/batch/group composition with dittoBarPlot

Cell metadata composition per / grouped by any other cell metadata can be
plotted with `dittoBarPlot()`

Some use cases:

- Assessing equivalence of batch representation across clusters

Primary inputs = object, var, group.by. (You can leave out those input names if
you like, as long as you provide the 'object' first, 'var' second, and
'group.by' third.)

```{r dittoBarPlot basic}
# Sex (or any metadata) breakdown within each cluster.
dittoBarPlot(object = full_sce, var = "sex", group.by = "cluster")

# Cluster make-up per pool / processing batch
#   Ideally, you would see relatively similar distributions here.
dittoBarPlot(full_sce, "cluster", "pool_id")
```

#### Some particlarly useful dittoBarPlot tweaks

```{r dittoBarPlot tweaks}
# Respect factor ordering using 'retain.factor.levels'
#   A 'flaw' relating to updates retaining backwards compatibility, the
#   developers plan to remove the need for this input in a future update.  But
#   as of the writing of this vignette, it is still needed.
dittoBarPlot(full_sce, "sex", group.by = "cluster_6x6",
    retain.factor.levels = TRUE
    )

# Ignore certain samples
#   Perhaps, as for the toy data set here, certain samples were only included as
#   a batch control, but do not have full information and were not intended for
#   inclusion in down-stream biological questions.
dittoBarPlot(full_sce, "sex", group.by = "cluster",
    cells.use = !full_sce$donor_id %in% c("HS2")
    )
```

### ScatterPlots, with all the same bells and whistles as umap plotting, dittoScatterPlot

Marker1 by marker2 expression-level scatter plots with dots (cells) optionally
colored by metadata or marker3 expression

Some use cases:

- Cluster annotation, sub-typing help. Ex: T cells CD45RA by CD27 or CCR7
- Imaging data, overlaying onto original imaging locations

Primary inputs = object, x.var, y.var (You can leave out those input names if
you like, as long as you provide the 'object' first, 'x.var' second, and
'y.var' third.)

```{r dittoScatterPlot basic}
dittoScatterPlot(object = full_sce, x.var = "CD45RA", y.var = "CCR7")

dittoScatterPlot(full_sce, "CD45RA", "CCR7",
    color.var = "sex")
```

#### Some particlarly useful dittoScatterPlot tweaks

```{r dittoScatterPlot tweaks}
# Adjust plotting order
#   Also try: "increasing" or "decreasing"
dittoScatterPlot(full_sce, "CD45RA", "CCR7", color.var = "sex",
    order = "randomize")

# Faceting (again, it's super useful!)
dittoScatterPlot(full_sce, "CD45RA", "CCR7", color.var = "cluster_6x6",
    split.by = "cluster_6x6")

# Setting titles & ...
#   The same 'main', 'sub', 'xlab', 'ylab' inputs can set titles across all
#   dittoSeq visualization functions (only exception is dittoHeatmap, the sole
#   non-ggplot function, where only 'main' can be used)
# focusing on only certain samples + cell types with cells.use in a case where perhaps the
#   control samples are particularly useful for deciding on meaningful value
#   cutoffs for the target markers.
dittoScatterPlot(full_sce, "CD45RA", "CCR7", color.var = "cluster_6x6",
    split.by = "cluster_6x6",
    cells.use = full_sce$control_sample & full_sce$coarse_annot == "CD4T")
```

### violin (or bar, or ridge) plots of marker expression across groups of cells with dittoPlot

Plots where marker expression-level per cell are plotted as violin and/or box
plots on a y-axes, or ridge plots in the x-axis direction.

Some use cases:

- Cluster annotation, sub-typing help. Ex: T cells CD45RA, CD27, or CCR7 alone across clusters

Primary inputs = object, var, group.by (You can leave out those input names if
you like, as long as you provide the 'object' first, 'var' second, and
'group.by' third.)

Very important additional input = plots, which sets the data representations to use. It defaults to c("jitter", "vlnplot") which puts a violin in front of jittered points for the individual cells, but changing to c("jitter", "vlnplot", "boxplot") will add boxplots on top. Additionally, wrappers **dittoBoxPlot**, **dittoRidgePlot**, and **dittoRidgeJitter** automatically adjust the plots input default to c("boxplot", "jitter"), c("ridgeplot"), and c("ridgeplot", "jitter"), respectively!

```{r dittoPlot basic}
dittoPlot(object = full_sce, var = "CD45RA", group.by = "cluster_6x6")

# For better examples, we'll focus on clusters 1:8 with cells.use
dittoPlot(object = full_sce, var = "CD45RA", group.by = "cluster_6x6",
    cells.use = full_sce$cluster %in% 1:8)

# Add a boxplot
dittoPlot(full_sce, "CD45RA", "cluster_6x6",
    cells.use = full_sce$cluster %in% 1:8,
    plots = c("jitter", "vlnplot", "boxplot"))

# With boxplot, but no violin plots
dittoPlot(full_sce, "CD45RA", "cluster_6x6",
    cells.use = full_sce$cluster %in% 1:8,
    plots = c("jitter", "boxplot"))
```

#### Some particlarly useful dittoPlot tweaks

```{r dittoPlot tweaks}
# Multiple genes in a single plot, by giving a set of markers to 'var'
dittoPlot(full_sce,
    var = c("CD45RA", "CD27", "CCR7"),
    group.by = "cluster",
    cells.use = full_sce$cluster %in% 1:8,
    # Facets will be used for the 'multivars'
    #   so we can set the faceting shape with split.ncol or split.nrow
    split.ncol = 1
    )

# Faceting or coloring as additional cell grouping mechanisms
dittoPlot(full_sce, "CD45RA",
    group.by = "sex",
    split.by = "coarse_annot")
dittoPlot(full_sce, "CD45RA",
    group.by = "coarse_annot",
    color.by = "sex",
    legend.title = "sex")
```

# 4. Run statistics on differences in cluster and cell type frequencies between samples.

Frequency comparison is a very common follow up in cytof data analysis that can
power identification of populations of interest.

The 'file_by_cluster_freq_norm' object output from cyclone can be used for
calculating stats, but this is only directly useful at the cluster-level.
Additional manipulation is required when wanting to calculate stats at the level
of coarse or fine cell annotations, which will often combine certain clusters.

For this reason, it can be easier to piggy-back off of dittoSeq's dittoFreqPlot
data collection in order to gather cell frequency numbers in a more flexible
way!

We've put together a function to help with this exact purpose, `freq_stats()`.

## freq_stats

`freq_stats()` is a function included in the `cyclone` package which flexibly
generates statistical comparisons between user-requested cell and sample
groupings by making use of `dittoFreqPlot()`'s data collection system.

### It's inputs are:

**Primary:**

- `object`: the SingleCellExperiment object to target (works for Seurat objects too, but we're not using that here). As with plotting, test with 'down_sce', but make sure to switch to 'full_sce' to make your final outputs!
- `group.by`: String. The name of a metadata within 'object' that holds the condition information you wish to compare between.
- `group.1` & `group.2`: Single values of the 'group.by' metadata which name the groups to compare.
- `sample.by`: String. The name of a metadata within 'object' that contains values which are unique for each sample. Typically, this can be "file_name" for cyclone outputs.
- `cell.by`: String. The name of a metadata within 'object' that contains the cluster or cell annotation information you wish to target.

**Secondary:**

- `cell.targs`: String vector, optional. If targeting just a subset of the cell clusters or annotations named in the 'cell.by' cell annotation metadata is desired, give that set of cell targets here.
- `cells.use`: Logical vector, optional. If targeting only a certain timepoint/treatment/condition is required for your biological question, *and your are using `group.by` for something else*, give this input `<object-name>$<condition-metadata-name> == '<target-condition>'` OR `<object-name>$<condition-metadata-name> %in% c('<target-condition1>', '<target-condition2>', '<target-condition3>')`
- `data.out`: Logical. FALSE by default. Setting it to TRUE will alter the output style to give a list of 2 elements: 'stats' = the standard output, and 'data' = the collected data.frame of cell counts and percentages used for the statistical calculations.
- `pseudocount`: Number, 0 by default (a.k.a. off by default), a small value to add, in fold_change calculations only, to both `group.1` and `group.2` median frequencies for the purpose of avoiding division by zero errors. When divide-by-zero errors are hit, we recommend using a value which is sufficiently smaller than expected cell fractions, such as 0.000001, in order to ensure that most other fold changes are only minimally affected. (Note: use of this pseudocount input has no effect on p-values.)
- `p.adjust.method`: String, "fdr" by default, which sets the type of p-value correction used. It's passed along to the `method` input of the base stats package `p.adjust` function, so any valid option for that function will work.

### How it works, briefly:

The function does 4 things:

1. Calls on dittoFreqPlot for data collection, frequency calculation, and percentage normalization.
2. Loops through each targeted cell type / annotation, calculating statistics by making us of R's `wilcox.test` function. This is a fitting statistical test for comparison of frequency data because, unlike a t-test, it does not assume the data has a normal distribution.
3. Combines outputs from all cell types / annotations into a single data.frame.
4. Applies a False Discovery Rate p-value adjustment.

To give a better understanding of the inputs to the function, additional details
on that first step can be helpful:

The function first makes a call to dittoFreqPlot to gather and normalize cell
count data. Within dittoFreqPlot, the number of cells of each 'sample.by' sample
assigned to each distinct 'cell.by' value are gathered. Cells' group information
is also pulled in at this stage, and only samples whose cells are marked as
'group.1' and 'group.2' in the 'group.by' metadata are targeted for counting.
The total number of cells for a given sample are then calculated (ignoring
any 'cell.targs' targeting, but respecting any 'cells.use' subsetting) and the
counts data is then normalized as percentages of all cells for the given sample.
Finally, if a set of cell names was given to 'cell.targs', this data.frame is
trimmed to only retain rows representing those 'cell_meta' values. At this
point, the data has been collected and normalized for the stats calculation
steps.

For additional details, check out the documentation with `?freq_stats`.

## Using freq_stats to compare between two groups

The frequency comparisons to run should be guided by your biological questions.

Was your study designed to assess differences between 2 groups? Then you want to target the metadata holding which cells belong to those 2 groups with `group.by` and give the names of those groups to `group.1` and `group.2`.

Was your study designed to assess difference between 3 groups? This function performs just pairwise comparisons, so you'll want to run it 3 times, targeting 1vs2, 1vs3, an 2vs3 to get the full picture of statistically significant differences. You likely have the group info in a single metadata, and if so you would keep `group.by` the same for all runs, while adjusting `group.1` and `group.2` for each of the distinct comparisons.

Another factor is what cell annotation level to target. Often, you'll want to assess all the levels you have. Here, we have two annotation levels, "coarse_annot", and "fine_annot", but we also have the individual clusters level as well. So that makes 3 cell-levels that we would target here.

Finally, is the question of whether you want the 'universe' that cell percentages are normalized within to be "all cells of the given sample" versus "all cells of a certain coarse annotation type". For example, you might be more interested in knowing the percentage of Treg cells out of CD4T cells, rather than the percentage of Treg cells out of all cells of the given sample. This can be achieved with our function as well!  Just skip down to the next section to see how!

**Here** we will compare across 'sex' which has only two groups in the data -> 1x pairwise group comparison. Plus, we have the "coarse_annot", "fine_annot", and "clusters" levels of cell annotation to compare between -> 3x cell type levels.  That makes for 3 comparisons total.

How this plays out in inputs to the function:

- `group.by`, `group.1`, `group.2`: In the data set for this tutorial, our metadata containing the sample grouping information is called 'sex', so we'll use that for 'group.by'. Its values are "F", representing female, and "M", representing Male, so we'll use 'group.1 = "F"' and 'group.2 = "M"'.
- `sample.by`: Typically, "file_name" can be used for this input for any cyclone data set unless you had multiple fcs files per sample.
- `cell.by`: Our two levels of annotations are stored in "coarse_annot" and "fine_annot" metadata columns, and we can also calculate statistics at the original "cluster" level, but we'll still want an idea of what those cells are, so we'll create a metadata column that's a combination of "fine_annot" and "cluster" below, and use that!

```{r coarse stats}
# Coarse level
freq_stats(
    object = full_sce,
    sample.by = "file_name",
    cell.by = "coarse_annot", 
    group.by = "sex", group.1 = "F", group.2 = "M")
```

```{r fine stats}
# Fine level
freq_stats(
    object = full_sce,
    sample.by = "file_name",
    cell.by = "fine_annot", 
    group.by = "sex", group.1 = "F", group.2 = "M")
```

```{r cluster stats}
### Cluster level, but with fine annotations + cluster names to make interpretation easier
# Create combined metadata
full_sce$fine_annot__cluster <- paste0(full_sce$fine_annot, "__cluster", full_sce$cluster)

# Run stats
freq_stats(
    object = full_sce,
    sample.by = "file_name",
    cell.by = "fine_annot__cluster", 
    group.by = "sex", group.1 = "F", group.2 = "M") # <-- Now using the metadata defined above
```

Of course, it's also helpful to have a visual.
We've described dittoFreqPlot in a previous section.
Here's how we might use it for the cluster-level:

```{r cluster viz, fig.width = 13, fig.height = 13}
# Visualization of frequency differences at the cluster-level
dittoFreqPlot(
    object = full_sce,
    var = "fine_annot__cluster",
    group.by = "sex",
    sample.by = "file_name",
    # Optionally target only group.1 and group.2 with:
    #   (modify "sex", "F", and "M" for your own data)
    cells.use = full_sce$sex %in% c("F", "M"),
    # Allow y-axis to shrink/expand per range of each cell type
    split.adjust = list(scale = "free_y")
)
```

## Using freq_stats to compare between two groups

One might also wish to assess cell frequency in terms of "fine" annotation per "coarse" annotation.  For example, you might be more interested in knowing the percentage of Treg cells out of CD4T cells, than in the percentage of Treg cells out of all cells of the given sample.  Luckily, to achieve such metrics, simply subset the SCE first!

```{r fine stats within CD4s}
# Create a subset of our data that is only the cells labeled as CD4T in coarse_annot
full_sce_CD4 <- full_sce[, full_sce$coarse_annot=="CD4T"]

# Calculate fine-level statistics within these CD4T cells
freq_stats(
    object = full_sce_CD4,
    sample.by = "file_name",
    cell.by = "fine_annot", 
    group.by = "sex", group.1 = "M", group.2 = "F",
    # Explicitly targeting just the values of 'fine_annot' here that have a 'coarse_annot' of CD4T 
    cell.targs = c("CD4T_EM", "CD4T_naive", "Tregs", "CD4T_CM")
)

# And the visualization
dittoFreqPlot(
    object = full_sce_CD4,
    var = "fine_annot",
    group.by = "sex",
    sample.by = "file_name",
    # Optionally target only group.1 and group.2 with:
    #   (modify "sex", "F", and "M" for your own data)
    cells.use = full_sce_CD4$sex %in% c("F", "M"),
    # vars.use is the cell.targs equivalent in the dittoFreqPlot function
    vars.use = c("CD4T_EM", "CD4T_naive", "Tregs", "CD4T_CM"),
    # Allow y-axis to shrink/expand per range of each cell type
    split.adjust = list(scale = "free_y"),
    # Update the y-axis label to mention the adjusted universe
    ylab = "Percent of CD4T cells"
)
```
